### 2014-11-05 motivation.tex
Major addition to implementation
Edits throughout

Consider how we might implement a highly available bank account on top of an
eventually consistent data store, with the \emph{integrity} constraint that the
balance should never drop below 0. We begin by implementing a bank
account replicated data type (RDT) in \name, and then describe the mechanisms
to obtain the desired correctness guarantees.

@@

A key novelty in \name is that it allows the addition of new RDTs to the store,
greatly improving the programmability of data stores. The programmer no
longer has to coerce the application logic to utilize the store provided data
types. In addition, \name treats the convergence semantics of the data type
separately from the consistency properties. This separation of concerns allows
the programmer to \emph{operationally} reason about conflict resolution, while
permitting \emph{declarative} reasoning for consistency. As mentioned in the
previous section, RDTs are described in terms of operations on a RDT object,
where each operation is described as a reduction over the state of the object
in terms of a list of effects. Let us assume that the bank account object provides
three operations: \cf{deposit}, \cf{withdraw} and \cf{getBalance}, with the
assumption that the withdraw fails if the account has insufficient balance.
Every operation in \name is of the following type, written in Haskell syntax:

\begin{codehaskell}
type Operation = (Effect e, Arg a, Res r)
               => [e] -> a -> (r, Maybe e)
\end{codehaskell}

\noindent It takes a list of effects (the \emph{context} for the operation),
and an input argument, and returns a result along with an optional effect. Read-only
operations do not return any effect. If an effect is returned, it is added to
the state of the object at the current replica, and asynchronously sent to
other replicas. The implementation of the bank account operations in \name is
given in Figure~\ref{fig:ex}:

@@

withdraw :: [Acc] -> Int -> (Bool, Maybe Acc)
withdraw ctxt v =
	if sel1 dollar getBalance ctxt () >= v
  then (True, Just dollar Withdraw v)
	else (False, Nothing)
\end{codehaskell}
\label{Definition of a bank account expressed in Quelea.}
\label{fig:ex}
\end{figure}

@@

Our goal is to choose the correct consistency level for each of the bank
account operations such that (1) the balance never drops below zero and (2) the
\cf{getBalance} operation never incorrectly returns a negative balance. To be
able to do this, we need to consider the anomalies that could arise under
eventual consistency.

@@

It is easy to see that the only way to ensure the integrity constraint on
\cf{balance} is to prevent concurrent withdrawals on the same bank account
object. This can be done by insisting that \cf{withdraw} be executed with
strong consistency properties. Although this prevents the integrity constraint
from being violated, a strongly consistent \cf{withdraw} itself is not
sufficient to prevent \cf{getBalance} from ever reporting a negative balance to
the end user.

Consider the execution shown in fig.~\ref{fig:negativeBalanceAnomaly}, which
consists of three concurrent sessions performing a \cf{deposit}, a
\cf{withdraw}, and a \cf{getBalance} operation, respectively, on the same bank
account object. As before, assume that the initial balance in the account is
zero. As the \cf{vis} edge indicates, operation \cf{withdraw(50)} in session 2,
can witnesses the effects of \cf{deposit(100)} from session 1, leading it to
rightfully conclude that the current balance is 50. Consequently,
\cf{withdraw(50)} is performed safely. The effect of this \cf{withdraw}
operation subsequently becomes visible to the \cf{getBalance} in Session 3.
However, because \cf{getBalance} only witnesses the effect of the \cf{withdraw}
operation from session 2, but not the \cf{deposit} from session 1, reports the
balance of negative 50 to the user.

Producing a negative balance is not the only inconsistency that the user is
exposed to under an eventually consistent bank account. Figure
~\ref{fig:missingUpdateAnomaly} shows an execution where the \cf{getBalance}
operation in a session does not witness the effects of an earlier \cf{withdraw}
operation performed in the same session, possibly because it was served by a
replica that has not yet merged the \cf{withdraw} effect, served by a different
 replica. This anomaly leads the user to incorrectly conclude that the
\cf{withdraw} operation failed to go through.

Although it is easy to understand the reasons behind the occurrence of the
aforementioned negative balance and missing update anomalies, finding the
appropriate fixes is not readily apparent. Making \cf{getBalance} a strongly
consistent operation is definitely sufficient to avert anomalies, but is it
really necessary? Given the cost of enforcing strong
consistency~\cite{DynamoDB, Pileus}, it is preferable to avoid it unless
there are no viable alternatives.  Exploring the space of these alternatives
requires understanding the subtle differences in semantics of various kinds
of weak consistency alternatives.

\subsection{Specifications and Contracts}

\name helps facilitate the mapping of operations to appropriate consistency
levels by letting the programmer declare application-level consistency
constraints as \emph{contracts} (Fig~\ref{fig:contract-lang}) that axiomatically
specify the set of allowed executions involving this operation. In the case of
the bank account, any execution that does not exhibit the anomalies
described in the previous section is a \emph{well-formed} execution on the bank
account object.  By specifying the set of legal executions for each data type in
terms of a trace of operation invocations on that type, \name\ ensures that all
executions over that type are well-formed.

In our running example, it is clear that in order to preserve the integrity
constraint, the \cf{withdraw} operation must be strongly consistent.  That is,
given two \cf{withdraw} operations $a$ and $b$, either $a$ is visible to $b$ or
vice versa. Utilizing the primitive relations described in
Section~\ref{sec:sysmod}, we express this application-level consistency
requirement as a contract or specification ($\cv_w$) over \cf{withdraw}:

@@

\noindent The expression $c:\rcf{deposit} \vee \rcf{withdraw}$ states that $c$ is an
effect that was emitted either by a \cf{deposit} or a \cf{withdraw} operation.
If a \cf{withdraw} $b$ is visible to \cf{getBalance}, then all \cf{deposit}
operations $a$ visible to the \cf{withdraw} should also be visible to
\cf{getBalance}. This prevents negative balance anomalies.

Our specification language provides operators to compose relations.  Thus,
the syntax $(R_1 \cap R_2)(a,b)$ is equivalent to $R_1(a,b) \wedge
R_2(a,b)$. The last line of the above contract says that if a \cf{deposit}
or a \cf{withdraw} operation precedes a \cf{getBalance} operation in session
order, and is applied on the same object as the \cf{getBalance} operation,
then it must be the case that the \cf{getBalance} operation witnesses the
effects of the preceding operations. Since there are no restrictions on when
or how a \cf{deposit} operation can execute, its contract is simply $\small
\true$.

@@

Notice that the contracts for \cf{withdraw} and \cf{getBalance} only express
application-level consistency requirements, and make no reference to the
semantics of the underlying store. To write contracts, a programmer only needs
to reason about the semantics of the application under the \name system model.
The mapping of application-level consistency requirements to appropriate
store-level guarantees is done automatically behind-the-scene.

How might one go about ensuring that an execution adheres to a contract? The
challenge is that a contract provides a declarative (axiomatic)
specification of an execution, while what is required is is an operational
procedure for \emph{enforcing} its implicit constraints. One strategy would
be to execute operations speculatively.  Here, operations are tentatively
applied as they are received from the client or other replicas. We can
maintain a runtime manifestation of executions, and check well-formedness
conditions at runtime, rolling back executions if they are
ill-formed. However, the overhead of state maintenance and the complexity of
user-defined contracts is likely to make this technique infeasible in
practice.

>>>>

Consider how we might implement a highly available bank account on top of an
eventually consistent data store, with the \emph{integrity} constraint that the
balance must be non-negative. We begin by implementing a bank account
replicated data type (RDT) in \name, and then describe the mechanisms to obtain
the desired correctness guarantees.

@@

A key novelty in \name is that it allows the addition of new RDTs to the store,
which obviates the need for coercing the application logic to utilize the store
provided data types. In addition, \name treats the convergence semantics of the
data type separately from its consistency properties. This separation of
concerns permits \emph{operational} reasoning for conflict resolution, and
\emph{declarative} reasoning for consistency. The combination of these
techniques enhances the programmability of the store.

Let us assume that the bank account object provides three operations:
\cf{deposit}, \cf{withdraw} and \cf{getBalance}, with the assumption that the
withdraw fails if the account has insufficient balance. Every operation in
\name is of the following type, written in Haskell syntax:

\begin{codehaskell}
type Operation = [e] -> a -> (r, Maybe e)
\end{codehaskell}

\noindent It takes a list of effects (the \emph{context} for the operation),
and an input argument, and returns a result along with an optional effect
(read-only operations return \cf{Nothing}). The new effect (if emitted) is
added to the state of the object at the current replica, and asynchronously
sent to other replicas. The implementation of the bank account operations in
\name is given in Figure~\ref{fig:ex}:

@@

withdraw :: [Acc] -> Int -> (Bool, Maybe Acc)
withdraw ctxt v =
	if sel1 dollar getBalance ctxt () >= v
  then (True, Just dollar Withdraw v)
	else (False, Nothing)
\end{codehaskell}
\caption{Definition of a bank account expressed in Quelea.}
\label{fig:ex}
\end{figure}

@@

Our goal is to choose the correct consistency level for each of the bank
account operations such that (1) the balance remains non-negative and (2) the
\cf{getBalance} operation never incorrectly returns a negative balance. Let us
first consider the anomalies that could arise under eventual consistency.

@@

It is easy to see that preventing concurrent \cf{withdraw} operations
eliminates this anamoly. This can be done by insisting that \cf{withdraw} be
executed as a strongly consistent operation. Despite this strengthening,
\cf{getBalance} operation may incorrectly report a negative balance to the end
user. Consider the execution shown in fig.~\ref{fig:negativeBalanceAnomaly},
which consists of three concurrent sessions performing a \cf{deposit}, a
\cf{withdraw}, and a \cf{getBalance} operation, respectively, on the same bank
account object. As the \cf{vis} edge indicates, operation \cf{withdraw(50)} in
session 2, witnesses the effects of \cf{deposit(100)} from session 1, concludes
that there is sufficient balance, and completes successfully. However, the
\cf{getBalance} operation may only witness this successful withdraw, but not
the causally preceding \cf{dopsit}, and reports the balance of negative 50 to
the user.

Under eventual consistency, the users may also be exposed to other forms of
inconsistencies. Figure~\ref{fig:missingUpdateAnomaly} shows an execution where
the \cf{getBalance} operation in a session does not witness the effects of an
earlier \cf{withdraw} operation performed in the same session, possibly because
it was served by a replica that has not yet merged the \cf{withdraw} effect.
This anomaly leads the user to incorrectly conclude that the \cf{withdraw}
operation failed to go through.

Although it is easy to understand the reasons behind the occurrence of the
aforementioned anomalies, finding the appropriate fixes is not readily
apparent. Making \cf{getBalance} a strongly consistent operation is definitely
sufficient to avert anomalies, but is it really necessary? Given the cost of
enforcing strong consistency~\cite{DynamoDB, Pileus}, it is preferable to avoid
it unless there are no viable alternatives. Exploring the space of these
alternatives requires understanding the subtle differences in semantics of
various kinds of weak consistency alternatives.

\subsection{Contracts}

\name helps facilitate the mapping of operations to appropriate consistency
levels by letting the programmer declare application-level consistency
constraints as \emph{contracts} (Fig~\ref{fig:contract-lang}\footnote{\name
exposes the contract construction language as a Haskell library}) that
axiomatically specify the set of allowed executions involving this operation.
In the case of the bank account, any execution that does not exhibit the
anomalies described in the previous section is a \emph{well-formed} execution
on the bank account object. By specifying the set of legal executions for each
data type in terms of a trace of operation invocations on that type, \name\
ensures that all executions over that type are well-formed.

In our running example, it is clear that in order to preserve the integrity
constraint, the \cf{withdraw} operation must be strongly consistent.  That is,
given two \cf{withdraw} operations $a$ and $b$, either $a$ is visible to $b$ or
vice versa. We express this application-level consistency requirement as a
contract ($\cv_w$) over \cf{withdraw}:

@@

\noindent The expression $c:\rcf{deposit} \vee \rcf{withdraw}$ states that $c$
is an effect that was emitted either by a \cf{deposit} or a \cf{withdraw}
operation. If a \cf{withdraw} $b$ is visible to \cf{getBalance} $\cureff$, then
all \cf{deposit} operations $a$ visible to $b$ should also be visible to
$\cureff$. This prevents negative balance anomalies. Our contract language
provides operators to compose relations. The syntax $(R_1 \cap R_2)(a,b)$ is
equivalent to $R_1(a,b) \wedge R_2(a,b)$. The last line of the above contract
says that if a \cf{deposit} or a \cf{withdraw} operation precedes a
\cf{getBalance} operation in session order, and is applied on the same object
as the \cf{getBalance} operation, then it must be the case that the
\cf{getBalance} operation witnesses the effects of the preceding operations.

Finally, since there are no restrictions on when or how a \cf{deposit}
operation can execute, its contract is simply $\small \true$.

@@

Notice that the contracts for \cf{withdraw} and \cf{getBalance} only express
application-level consistency requirements, and make no reference to the
semantics of the underlying store. To write contracts, a programmer only needs
to reason about the semantics of the application under the \name system model.
The mapping of application-level consistency requirements to appropriate
store-level guarantees is done automatically behind-the-scene. How might one go
about ensuring that an execution adheres to a contract? The challenge is that a
contract provides a declarative (axiomatic) specification of an execution,
while what is required is an operational procedure for \emph{enforcing} its
implicit constraints.

One strategy would be to execute operations speculatively.  Here, operations
are tentatively applied as they are received from the client or other replicas.
We can maintain a runtime manifestation of executions, and check
well-formedness conditions at runtime, rolling back executions if they are
ill-formed. However, the overhead of state maintenance and the complexity of
user-defined contracts is likely to make this technique infeasible in practice.
